---
title: "Decoding Wine Quality"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
runtime: shiny
---

```{r setup, include=FALSE}
# Import dataset (wine_dataset) from an .R document:
source ("dataset_dashboard.R")
# Libraries
library(tidyverse)
library(shiny)
library(shinydashboard)
library(flexdashboard)
library(plotly)
```

Variables Correlations
=======================================================================

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
selectInput("type", label = "Select wine type:",
            choices = unique(wine_dataset$type), selected = "red")
```
\

**Variables that influence both red and white wines:**

<span style="font-size: 11px;">
**Fixed acidity:** influences taste, pH, and overall wine life.
\
**Volatile acidity:** refers to steam-distillable acids.
\
**Citric acid:** provides a certain freshness to the wine.
\
**Residual sugar:** the amount of sugar that could not be fermented.
\
**Chlorides:** mineral salts in the wine.
\
**Free sulfur dioxide:** a powerful antimicrobial agent.
\
**Total sulfur dioxide:** used as a preservative, antioxidant, and antiseptic.
\
**Density:** also known as "specific gravity," an analytical parameter of wine.
\
**pH:** a measure of acidity or alkalinity in an aqueous solution.
\
**Sulphates:** another preservative, antimicrobial, and antioxidant.
\
**Alcohol:** being wine a fermented beverage, it has an alcohol content.
\
**Quality:** a rating reflecting the overall quality of the wine.
</span>

Row
-----------------------------------------------------------------------

### Dataset records {.value-box}
```{r}
valueBox(value = nrow(wine_dataset), "Dataset records", icon = "fa-database")
```

### Wine type records {.value-box}
```{r}
type_sample <- reactive({
  filter(wine_dataset, type == input$type)
})

renderValueBox({
  valueBox(value = nrow(type_sample()), "Wine type records", icon = "fa-wine-glass")
})
```

### Average quality {.value-box}
```{r}
avg_quality <- reactive({
  filter(wine_dataset, type == input$type)
})

renderValueBox({
  valueBox(value = round(mean(avg_quality()$quality), 1), "Average quality", icon = "fa-circle-check")
})
```

Row
-----------------------------------------------------------------------

### **Correlation heatmap** {data-width=650}
```{r}
renderPlotly({
  selected_cor <- switch(input$type, "red" = red_cor, "white" = white_cor, red_cor)
    
plot_ly(x = colnames(selected_cor), y = rownames(selected_cor), z = selected_cor, type = "heatmap")
  })
```

### **Quality correlation and p-value** {data-width=370}
<span style="font-size: 11px;">
_Statistically significant if p_value < 0.05_
</span>
```{r}
renderTable({
  corr_pvalue <- switch(input$type, "red" = df_red_cor, "white" = df_white_cor, df_red_cor)
  
  corr_pvalue}
  , digits = 2, container = tags$table(style = 'overflow-x:auto;'), rownames = TRUE)
```

Variables Analysis
=======================================================================

Column {.sidebar}
-----------------------------------------------------------------------

```{r}
sliderInput("quality", label = "Select wine quality:",
            min = 1, max = 9, value = 5, step = 1)
```
**Sample size:**
```{r}
quality_sample <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderText({nrow(quality_sample())})
```
\

**Variables that influence wines' quality:**

<span style="font-size: 11px;">
**Fixed acidity:** influences taste, pH, and overall wine life.
\
**Volatile acidity:** refers to steam-distillable acids.
\
**Citric acid:** provides a certain freshness to the wine.
\
**Residual sugar:** the amount of sugar that could not be fermented.
\
**Chlorides:** mineral salts in the wine.
\
**Free sulfur dioxide:** a powerful antimicrobial agent.
\
**Total sulfur dioxide:** used as a preservative, antioxidant, and antiseptic.
\
**Density:** also known as "specific gravity," an analytical parameter of wine.
\
**pH:** a measure of acidity or alkalinity in an aqueous solution.
\
**Sulphates:** another preservative, antimicrobial, and antioxidant.
\
**Alcohol:** being wine a fermented beverage, it has an alcohol content.
</span>

Row
-----------------------------------------------------------------------

### **Fixed acidity**
```{r}
fixedac <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(fixedac(), y = ~fixed.acidity, type = "box", 
            color = ~type, colors = c("#8E0000", "#E7D40A")) %>%
    layout(yaxis = list(title = "g/dm3"))
  })
```

### **Citric acid**
```{r}
citrica <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(citrica(), y = ~citric.acid, type = "box", 
            color = ~type, colors = c("#8E0000", "#E7D40A")) %>%
    layout(yaxis = list(title = "g/dm3"))
  })
```

### **Volatile acidity**
```{r}
volatile <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(volatile(), y = ~volatile.acidity, type = "box", 
            color = ~type, colors = c("#8E0000", "#E7D40A")) %>%
    layout(yaxis = list(title = "g/dm3"))
  })
```

Row
-----------------------------------------------------------------------

### **PH**
```{r}
ph <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(ph(), x = ~pH, type = "histogram", 
            color = ~type, colors = c("#8E0000", "#E7D40A")) %>%
    layout(xaxis = list(range = c(2.7, 3.8)))
  })
```

### **Density**
```{r}
density <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(density(), x = ~type, y = ~density, split = ~type, type = "violin",
            box = list(visible = T), meanline = list(visible = T), 
            color = ~type, colors = c("#8E0000", "#E7D40A")) %>%
    layout(yaxis = list(title = "Density (g/cm3)"))
  })
```

### **Chlorides vs sulphates**
```{r}
chlo_sulph <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(chlo_sulph(), x = ~sulphates, y = ~chlorides, type = "scatter", 
          mode = "markers", color = ~type, colors = c("#8E0000", "#E7D40A"),
          opacity = 0.7) %>%
    layout(xaxis = list(title = "Sulphates (g/dm3)", range = c(0.26, 2)),
           yaxis = list(title = "Chlorides (g/dm3)"))
  })
```

Row
-----------------------------------------------------------------------

### **Alcohol vs residual sugar**
```{r}
alcohol_sugar <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(alcohol_sugar(), x = ~residual.sugar, y = ~alcohol, type = "scatter", 
          mode = "markers", color = ~type, colors = c("#8E0000", "#E7D40A"),
          opacity = 0.7) %>%
    layout(xaxis = list(title = "Residual Sugar (g/dm^3)"), yaxis = list(title = "Alcohol (%)"))
  })
```

### **Sulfur dioxide**
```{r}
sulfur_dioxide <- reactive({
  filter(wine_dataset, quality == input$quality)
})

renderPlotly({
    plot_ly(sulfur_dioxide(), x = ~total.sulfur.dioxide, y = ~free.sulfur.dioxide, type = "scatter", 
            mode = "markers", color = ~type, colors = c("#8E0000", "#E7D40A"),
            opacity = 0.7) %>%
    layout(xaxis = list(title = "Total (mg/dm^3)"), yaxis = list(title = "Free (mg/dm^3)"))
  })
```